Modelling palaeoecological time series using generalized additive models

### 2018-05-14 manuscript.tex
submitted version of the manuscript

The vast majority of data produced by palaeoecologists and
palaeolimnologists is in the form of time-ordered observations on one or
more proxies or biological taxa (Birks, 2012b; Smol, 2008; Smol et al.,
2012). Typically these data are arranged irregularly in time; in the
absence of annual laminae or varves, the sediment core is sectioned at
regular depth intervals (Glew et al., 2001), which, owing to variation
in accumulation rates over time and compaction by overlying sediments,
results in an uneven sampling in time. An under-appreciated secondary
feature of such sampling is that younger sediments often have larger
variance than older sediments; each section of core represents fewer
lake years in newer samples, relative to older samples. This variable
time averaging acts as a variable low-pass (high-cut) filter of the raw,
annual depositional signal.

The irregular intervals between samples means that the customary
time-series analysis methods of autoregressive or moving average
processes, in the form of autoregressive integrated moving average
(ARIMA) models, are practically impossible to apply because software
implementing these methods typically requires even spacing of
observations in time. Dutilleul et al. (2012) and Birks (2012a),
eschewing the term \emph{time series}, prefer to call such data
\emph{temporal series} on account of the irregular spacing of samples, a
distinction that I find unnecessary.

Where statistical approaches have been applied to trend estimation in
palaeoenvironmental time series, the most commonly used has been
\loess{} (Birks, 1998, 2012a; Cleveland, 1979; Juggins and Telford,
2012). \loess{}, or locally weighted scatterplot smoother, as it's name
suggests, was developed to smooth x-y scatterplot data, ostensibly as
part of exploratory data analysis (Cleveland, 1979). The method fits a
smooth line through data by fitting weighted least squares (WLS) models
to observations within a particular, user-specified window of the target
point, whose width is typically expressed as a proportion \(\alpha\) of
the \(n\) data points. Weights are determined by how close (in the
x-axis only) an observation in the window is to the focal point, with a
tricubic function, by default, giving greatest weight to points closest
to the focal point. The interim \loess{}-smoothed value for the focal
point is the predicted value from the weighted regression at the focal
point. The interim values are updated using weights based on how far in
the y-axis direction, the interim smoothed value lies from the observed
value plus the x-axis distance weights; this has the effect of
down-weighting the effects of outlier observations. The final \loess{}
is obtained by joining up the smoothed (fitted) values. The user has to
choose how large a window to use, whether to fit degree 1 (linear) or
degree 2 (quadratic) polynomials in the WLS model, and how to weight
points in the x-axis. When used in an exploratory mode, the user has
considerable freedom to choose the detail of the \loess{} fit; the
window width, for example, can be infinitely tweaked to give as close a
fit to the data, as assessed by eye, as is desired. Using
cross-validation (CV) to choose \(\alpha\) or the degree of polynomial
in the WLS model is complicated for a number of reasons, not least that
the CV scheme used must involve the time ordering of the data (e.g.
Bergmeir et al., 2018). This level of subjectivity is problematic
however once we wish to move beyond exploratory data analysis to
statistically identify trends and to test hypotheses involving those
trend estimates.

Running means or other types of filter (Juggins and Telford, 2012) have
also been used extensively to smooth palaeoenvironmental time series,
but, as with \loess{}, their behaviour depends on a number of factors,
including the filter width. Furthermore, the width of the filter causes
boundary issues; with a centred filter, of width five data points, the
filtered time series would be two data points shorter at both ends of
the series because the filter weights are not defined for the first and
last two observations of the original series as these time points were
not observed. Considerable research effort has been expended to identify
ways to pad the original time series at one or both ends to maintain the
original length in the filtered series, without introducing bias because
of the method of padding (e.g. Mann, 2004, 2008; Mills, 2006, 2007,
2010).

These are by no means the only methods that have been used to estimated
trends in stratigraphic series. Another common approach involves fitting
a simple linear trend using ordinary least squares regression and use
the resulting \emph{t} statistic as evidence against the null hypothesis
of no trend despite the statistical assumptions of the method being
almost surely violated due to dependence among observations arising from
temporal autocorrelation in the series. The Pearson correlation
coefficient, \emph{r}, is also often used to detect trends in palaeo
time series (Birks, 2012a), despite the fact that \emph{r} provides no
information as to the magnitude of the estimated trend, and the same
temporal autocorrelation problem that dogs ordinary least squares
similarly plagues significance testing for \emph{r} (Tian et al., 2011).
Additionally, both the simple least squares trend line and \emph{r} are
tests for \emph{linear} trends only, and yet we typically face data sets
with potentially far more complex trends than can be identified by these
methods. Instead, non-parametric rank correlation coefficients have been
used (Birks, 2012a; Gautheir, 2001), and whilst these do allow for the
detection of non-linear trends, trends are restricted to be monotonic,
no magnitude of the trend is provided, and the theory underlying
significance testing of Spearman's \(\rho\) and Kendall's \(\tau\)
assumes independent observations.

Palaeoecology and palaeolimnology have moved away from being descriptive
disciplines, rapidly adopting new statistical developments in the 1990s
and beyond (@ Smol et al., 2012). Less development has been observed in
the area of trend estimation in palaeoenvironmental time series. Here, I
describe the use of generalized additive models (GAMs; Hastie and
Tibshirani, 1986, 1990; Ruppert et al., 2003; Wood, 2017; Yee and
Mitchell, 1991) for trend estimation. GAMs, like simple linear
regression, are a formal regression-based method for estimating trends,
yet they are also, superficially at least, similar to \loess{}. GAMs and
\loess{} estimate smooth, non-linear trends in time series and both can
handle the irregular spacing of samples in time, yet GAMs do not suffer
from the subjectivity that plagues \loess{} as a method of formal
statistical inference.

In the subsequent sections, I present an introduction to GAMs and how
they are able to estimate smooth, non-linear trends from time series.
Next the issue of uncertainty in model-estimated trends is discussed, as
is the topic of posterior simulation from a regression model. The issue
of identifying periods of significant environmental change is considered
and I show how this can be achieved using the first derivative of the
estimated trend. Two non-standard types of spline --- adaptive smoothers
and Gaussian process splines --- that are especially applicable to GAMs
in the palaeoenvironmental setting are subsequently described, followed
by an assessment of the the impact of age-model uncertainty on trend
estimation via GAMs. Finally, I briefly discuss the application of GAM
trend analysis to multivariate species abundance and compositional data.

@@

To illustrate the estimation of trends in palaeoenvironmental data using
GAMs, I use two proxy time series; a 150-year bulk organic matter
δ\textsuperscript{15}N time series from Small Water, and a 3000-year
alkenone record from Braya-Sø. Between them, the two examples, combine
many of the features of interest to palaeoecologists that motivate the
use of GAMs; non-linear trends and the question of when changes in the
measured proxy occurred. The example analyses were all performed using
the \emph{mgcv} package and R, and the supplementary material contains a
fully annotated document showing the R code used to replicate all the
analyses described in the remainder of the paper.

@@

where \(\beta_0\) is a constant term, the model \emph{intercept},
representing the expected value of \(y_t\) where \(x_t\) is 0.
\(\beta_1\) is the \emph{slope} of the best fit line through the data;
it measures the rate of change in \(y\) for a unit increase in \(x\).
The unknowns, the \(\beta_j\) are commonly estimated using least squares
by minimising the sum of squared errors, \(\sum_t \varepsilon_t^2\). If
we want to ask if the estimated trend \(\beta_1\) is statistically
significant, a process called \emph{inference}, we must make further
assumptions about the data (conditional upon the fitted model) or the
model errors (residuals);
\(\varepsilon_t \stackrel{iid}{\sim} \mathcal{N}(0, \sigma^2)\). This
notation indicates that the residuals \(\varepsilon_t\) are
\emph{independent} and \emph{identically distributed} Gaussian random
variables with mean equal to \(0\) and constant variance \(\sigma^2\).
In the time series setting, the assumption of independence of model
residuals is often violated.

The linear model described above is quite restrictive in terms of the
types of trend it can fit; essentially linear increasing or decreasing
trends, or, trivially, a null trend of no change. This model can be
extended to allow for non-linear trends, most notably by making \(y_t\)
depend on polynomials of \(x_t\), for example

@@

where polynomials of \(x_t\) up to order \(P\) are used. This model
allows for more complex trends but it remains a fully parametric model
and suffers from several problems, especially the behaviour of the
fitted polynomial at the start and end of the observed series.

@@

Can we do better than these polynomial fits? In the remainder, I hope to
demonstrate that the answer to that question is emphatically ``yes''!
Below I describe a coherent and consistent approach to modelling
palaeoenvironmental time series using generalized additive models, which
builds upon the linear regression framework.

@@

where the linear effect of time(the \(\beta_1 x_t\) part) has been
replaced by a smooth function of time, \(f(x_t)\). The immediate
advantage of the GAM is that we are no longer restricted to the shapes
of trends that can be fitted via global polynomial functions such as
\eqref{eq:polynomial-model}. Instead, the shape of the fitted trend will
be estimated from the data itself.

The linear model is a special case of a broader class know as the
generalized linear model (GLM; McCullagh and Nelder, 1989). The GLM
provides a common framework for modelling a wide range of types of data,
such as count, proportions, or binary (presence/absence) data, that are
not conditionally distributed Gaussian. GLMs are, like the linear model,
parametric in nature; the types of trends that we can fit using a GLM
are the linear or polynomial models. GAMs extend the GLM by relaxing
this parametric assumption; in a GAM some, or all, of the parametric
terms, the \(\beta_p\), are replace by smooth functions \(f_j\) of the
covariates \(x_j\). For completeness then, we can write
\eqref{eq:additive-model} as a GLM/GAM

@@

where \(\mu_t\) is the expected value (e.g.~the mean count or the
probability of occurrence) of the random variable \(Y_t\)
(\(\mu_t \equiv \mathbb{E}(Y_t)\)) of which we have observations
\(y_t\). \(g\) is the link function, an invertible, monotonic function,
such as the natural logarithm, and \(g^{-1}\) is its inverse. The link
function maps values from the response scale on to the scale of the
linear predictor, whilst the inverse of thew link function provides the
reverse mapping. For example, count data are strictly non-negative
integer values and are commonly modelled as a Poisson GLM/GAM using the
natural log link function. On the log scale, the response can take any
real value between \(-\infty\) and \(+\infty\), and it is on this scale
that model fitting actually occurs (i.e.~using equation
\eqref{eq:gam-linear-pred}). However we need to map these unbounded
values back on to the non-negative response scale. The inverse of the
log link function, the exponential function, achieves this and maps
values to the interval 0--\(\infty\) (equation \eqref{eq:gam-inverse}).

@@

It is clear from the plots of the data that we require the fitted trends
for the Small Water δ\textsuperscript{15}N and Braya-Sø \uk{} time
series to be non-linear functions, but it is less clear how to specify
the actual shape require. Ideally, we'd like to learn the shape of the
trend from the data themselves. We will refer to these non-linear
functions as \emph{smooth functions}, or just \emph{smooths} for short,
and we will denote a smooth using \(f(x_t)\). Further, we would like to
represent the smooths in a way that \eqref{eq:gam} is represented
parametrically so that it can be estimate within the well-studied GLM
framework. This is achieved by representing the smooth using a
\emph{basis}. A basis is a set of functions that collectively span a
space of smooths that, we hope, contains the true \(f(x_t)\) or a close
approximation to it. The functions in the basis are known as \emph{basis
functions}, and arise from a \emph{basis expansion} of a covariate.
Writing \(b_j(x_t)\) as the \(j\)th basis function of \(x_t\), the
smooth \(f(x_t)\) can be represented as a weighted sum of basis
functions

@@

where \(\beta_j\) is the weight applied to the \(j\)th basis function.
Note that here we concern ourselves only with univariate smooths
involving a single covariate, the time variable \(x_t\).

The polynomial model we encountered earlier is an example of a
statistical model that uses a basis expansion. For the cubic polynomial
(\(P = 3\)) fit shown in Figure \ref{fig:polynomial-example-plot} there
are in fact 4 basis functions: \(b_1(x_t) = x_t^0 = 1\),
\(b_2(x_t) = x_t\), \(b_3(x_t) = x^2_t\), and \(b_4(x_t) = x_t^3\). Note
that \(b_1(x_t)\) is constant and is linked to the model intercept,
\(\beta_0\), in the linear model \eqref{eq:polynomial-model}, and
further, that the weights are the estimated coefficients in the model,
the \(\beta_j\).

@@

A cubic spline is a smooth curve comprised of sections of cubic
polynomials (\(P = 3\)), where the sections are joined together at some
specified locations --- known as \emph{knots} --- in such a way that at
the joins, the two sections of cubic polynomial that meet have the same
value as well as the same first and second derivative. These properties
mean that the sections join smoothly and differentiably at the knots
(Wood, 2017, 5.3.1).

@@

To estimate a model using this basis expansion each basis function forms
a column in the model matrix \(\boldsymbol{X}\) and the weights
\(\beta_j\) can be found using least squares regression (assuming a
Gaussian response). Note that in order to estimate a coefficient for
each basis function the model has to be fitted without an intercept
term. In practice we would include an intercept term in the model and
therefore the basis functions are modified via an identifiability
constraint. This has the effect of making the basis orthogonal to the
intercept but results in more complicated basis functions than those
shown in in Figure \ref{fig:basis-function-example-plot}a.

@@

Cubic regression splines, as well as many other types of spline, require
the analyst to choose the number and location of the knots that
parametrise the basis. Thin plate regression splines (TPRS) remove this
element of subjectivity when fitting GAMs. Thin plate splines were
introduced by Duchon (1977) and, as well as solving the knot selection
problem, have several additional attractive properties in terms of
optimality and their ability to estimate a smooth function of two or
more variables, leading to smooth interactions between covariates.
However, thin plate splines have one key disadvantage over CRS; thin
plate splines have as many unknown parameters as there are unique
combinations of covariate values in a data set (Wood, 2017, 5.5.1). It
is unlikely in the extreme that any real data problem would involve
functions of such complexity that they require as many basis functions
as data. It is much more likely that the true functions that we attempt
to estimate are far simpler than the set of functions representable by 1
basis function per unique data value. From a practical point of view, it
is also highly inefficient to carry around all these basis functions
whilst model fitting, and the available computational resources would
become quickly exhausted for large time series with many observations.

To address this issue, thin plate regression splines (TPRS) have been
suggested which truncate the space of the thin plate spline basis to
some lower number of basis functions whilst preserving much of the
advantage of the original basis as an optimally-fitting spline (Wood,
2003). A rank 7 TPRS basis (i.e.~one containing 7 basis functions) is
shown in Figure \ref{fig:basis-function-example-plot}c for the Small
Water time series. The truncation is achieved by performing an
eigen-decomposition of the basis functions and retaining the
eigenvectors associated with the \(k\) largest eigenvalues. This is
similar to the way principal components analysis decomposes a data set
into axes of variation (eigenvectors) in decreasing order of variance
explained. The truncated basis can preserve much of the space of
functions spanned by the original basis but at the cost of using far
fewer basis functions (Wood, 2003, 2017, 5.5.1). Note the horizontal
TPRS basis function (at δ\textsuperscript{15}N = 1) in Figure
\ref{fig:basis-function-example-plot}c; this basis function is
confounded with the intercept term and, after the application of
identifiability constraints, ends up being removed from the set of basis
functions used to fit the model.

The truncation suggested by Wood (2003) is not without cost; the
eigen-decomposition and related steps can be relatively costly for large
data sets. For data sets of similar size to the two examples used here,
the additional computational effort required to set up the TPRS basis
over the CRS basis will not be noticeable. For highly resolved series
containing more than \textasciitilde{}1000 observations the truncation
may be too costly computationally. In such instances, little is lost by
moving to the CRS basis, with the same number of knots as the rank of
the desired TPRS, with the benefit of considerably reduced set up time
for the basis.

@@

In practice, the size of the basis is an upper limit on the expected
complexity of the trend, and a simple test is available to check if the
basis used was sufficiently large (Pya and Wood, 2016). This test is
available via the \texttt{gam.check()} function in \emph{mgcv} for
example, which essentially looks at whether there is any additional
nonlinearity or structure in the residuals that can be explained by a
further smooth of \(x_t\). Should a smooth term in the fitted model fail
this test the model can be refitted using a larger basis expansion, say
by doubling the value of \emph{k} (the rank) used to fit the original.
Note also that a smooth might fail this test whilst using fewer
effective degrees of freedom than the maximum possible for the dimension
of basis used. This may happen when the true function lies at the upper
limit of the set of functions encompassed by the size of basis used.
Additionally, a basis of size \(2k\) encompasses a richer space of
functions of a given complexity than a basis of size \(k\) (Wood, 2017);
increasing the basis dimension used to fit the model may unlock this
additional function space resulting in a better fitting model whilst
using a similar number of effective degrees of freedom.

@@

Having identified regression splines as a useful way to represent \(f\),
we next need a way to decide how wiggly the fitted trend should be. A
backwards elimination approach to sequentially remove knots or basis
functions might seem appropriate, however such an approach would likely
fail as the resulting sequence of models would not be strictly nested,
precluding many forms of statistical comparison (Wood, 2017).
Alternatively, we could keep the basis dimension at a fixed size but
guard against fitting very complex models through the use of a
wiggliness penalty.

@@

The trend in δ\textsuperscript{15}N values is clearly non-linear and it
would be difficult to suggest a suitable polynomial model that would
allow for periods of relatively no change in δ\textsuperscript{15}N as
well as rapid change. Instead, a GAM is ideally suited to modelling such
trends; the data suggest a smoothly varying change in
δ\textsuperscript{15}N between 1925 and 1975. It is reasonable to expect
some autocorrelation in the model errors about the fitted trend.
Therefore I fitted the following GAM to the δ\textsuperscript{15}N time
series.

@@

Fitting this model as a standard GAM with REML smoothness selection
resulted in the same fitted trend as the GAM with CAR(1) errors (not
shown), whilst using GCV smoothness selection resulted in a much more
satisfactory fitted trend. There are two potential problems with the
GCV-selected trend: i) GCV is sensitive to the profile of the GCV score
and has been shown to badly under smooth data in situations where the
profile is flat around the minimum GCV score, and ii) the model fitted
assumes that the observations are independent, an assumption that is
certainly violated in the \uk{} time series.

>>>>

Palaeoecology and palaeolimnology have moved away from being descriptive
disciplines, rapidly adopting new statistical developments in the 1990s
and beyond (@ Smol et al., 2012). Less development has been observed in
the area of trend estimation in palaeoenvironmental time series. The
vast majority of data produced by palaeoecologists and
palaeolimnologists is in the form of time-ordered observations on one or
more proxies or biological taxa (Birks, 2012b; Smol, 2008; Smol et al.,
2012). Typically these data are arranged irregularly in time; in the
absence of annual laminae or varves, the sediment core is sectioned at
regular depth intervals (Glew et al., 2001), which, owing to variation
in accumulation rates over time and compaction by overlying sediments,
results in an uneven sampling in time. An under-appreciated feature of
such sampling is that younger sediments often have larger variance than
older sediments; each section of core represents fewer lake years in
newer samples, relative to older samples. This variable averaging acts
as a time-varying low-pass (high-cut) filter of the annual depositional
signal.

Irregular intervals between samples means that the time-series analysis
methods of autoregressive or moving average processes, in the form of
autoregressive integrated moving average (ARIMA) models, are practically
impossible to apply because software typically requires even spacing of
observations in time. Dutilleul et al. (2012) and Birks (2012a),
eschewing the term \emph{time series}, prefer to call such data
\emph{temporal series} on account of the irregular spacing of samples, a
distinction that I find unnecessary.

Where statistical approaches have been applied to trend estimation in
palaeoenvironmental time series, a commonly-used method is \loess{}
(Birks, 1998, 2012a; Cleveland, 1979; Juggins and Telford, 2012).
\loess{}, locally weighted scatterplot smoother, as it's name suggests,
was developed to smooth x-y scatterplot data (Cleveland, 1979). The
method fits a smooth line through data by fitting weighted least squares
(WLS) models to observations within a particular, user-specified window
of the focal point, whose width is typically expressed as a proportion
\(\alpha\) of the \(n\) data points. Weights are determined by how close
(in the x-axis only) an observation in the window is to the focal point
giving greatest weight given to points closest to the focal point. The
interim \loess{}-smoothed value for the focal point is the predicted
value from the weighted regression at the focal point. The interim
values are updated using weights based on how far in the y-axis
direction the interim smoothed value lies from the observed value plus
the x-axis distance weights; this has the effect of down-weighting
outlier observations. The final \loess{} is obtained by joining the
smoothed values. The user has to choose how large a window to use,
whether to fit degree 1 (linear) or degree 2 (quadratic) polynomials in
the WLS model, and how to weight points in the x-axis. When used in an
exploratory mode, the user has considerable freedom to choose the detail
of the \loess{} fit; the window width, for example, can be infinitely
tweaked to give as close a fit to the data, as assessed by eye, as is
desired. Using cross-validation (CV) to choose \(\alpha\) or the degree
of polynomial in the WLS model is complicated for a number of reasons,
not least because the CV scheme used must involve the time ordering of
the data (e.g. Bergmeir et al., 2018). This subjectivity is problematic
however once we wish to move beyond exploratory analysis and
statistically identify trends to test hypotheses involving those trend
estimates.

Running means or other types of filter (Juggins and Telford, 2012) have
also been used extensively to smooth palaeoenvironmental time series,
but, as with \loess{}, their behaviour depends on a number of factors,
including the filter width. Furthermore, the width of the filter causes
boundary issues; with a centred filter, of width five, the filtered time
series would be two data points shorter at both ends of the series
because the filter values are not defined for the first and last two
observations of the original series as these extra time points were not
observed. Considerable research effort has been expended to identify
ways to pad the original time series at one or both ends to maintain the
original length in the filtered series, without introducing bias due to
the padding (e.g. Mann, 2004, 2008; Mills, 2006, 2007, 2010).

These are not the only methods that have been used to estimated trends
in stratigraphic series. Another common approach involves fitting a
simple linear trend using ordinary least squares regression and use the
resulting \emph{t} statistic as evidence against the null hypothesis of
no trend despite the statistical assumptions being almost surely
violated due to dependence among observations. The Pearson correlation
coefficient, \emph{r}, is also often used to detect trends in palaeo
time series (Birks, 2012a), despite the fact that \emph{r} provides no
information as to the magnitude of the estimated trend, and the same
temporal autocorrelation problem that dogs ordinary least squares
similarly plagues significance testing for \emph{r} (Tian et al., 2011).
Additionally, both the simple least squares trend line and \emph{r} are
tests for \emph{linear} trends only, and yet we typically face data sets
with potentially far more complex trends than can be identified by these
methods. Instead, non-parametric rank correlation coefficients have been
used (Birks, 2012a; Gautheir, 2001), and whilst these do allow for the
detection of non-linear trends, trends are restricted to be monotonic,
no magnitude of the trend is provided, and the theory underlying
significance testing of Spearman's \(\rho\) and Kendall's \(\tau\)
assumes independent observations.

Here, I describe generalized additive models (GAMs; Hastie and
Tibshirani, 1986, 1990; Ruppert et al., 2003; Wood, 2017; Yee and
Mitchell, 1991) for trend estimation. GAMs, like simple linear
regression, are a regression-based method for estimating trends, yet
they are also, superficially at least, similar to \loess{}. GAMs and
\loess{} estimate smooth, non-linear trends in time series and both can
handle the irregular spacing of samples in time, yet GAMs do not suffer
from the subjectivity that plagues \loess{} as a method of formal
statistical inference.

In the subsequent sections, I present an introduction to GAMs and
discuss the issue of uncertainty in model-estimated trends, the topic of
posterior simulation from a regression model and how to identify periods
of significant environmental change using the first derivative of the
estimated trend. Two non-standard types of spline --- adaptive smoothers
and Gaussian process splines --- that are especially applicable to GAMs
in the palaeoenvironmental setting are subsequently described, followed
by an assessment of the the impact of age-model uncertainty on trend
estimation via GAMs. Finally, I briefly discuss the application of GAM
trend analysis to multivariate species abundance and compositional data.

@@

To illustrate trend estimation in palaeoenvironmental data using GAMs, I
use two proxy time series; a 150-year bulk organic matter
δ\textsuperscript{15}N record from Small Water, and a 3000-year alkenone
record from Braya-Sø. Between them, the two examples, combine many of
the features of interest to palaeoecologists that motivate the use of
GAMs; non-linear trends and the question of when changes in the measured
proxy occurred. The example analyses were all performed using the
\emph{mgcv} package (version 1.8.23; Wood, 2017) and R (version 3.4.4; R
Core Team, 2018), and the supplementary material contains a fully
annotated document showing the R code used to replicate all the analyses
described in the remainder of the paper.

@@

where \(\beta_0\) is a constant term, the model \emph{intercept},
representing the expected value of \(y_t\) where \(x_t\) is 0.
\(\beta_1\) is the \emph{slope} of the best fit line through the data;
it measures the rate of change in \(y\) for a unit increase in \(x\).
The unknowns, the \(\beta_j\), are commonly estimated using least
squares by minimising the sum of squared errors,
\(\sum_t \varepsilon_t^2\). If we want to ask if the estimated trend
\(\beta_1\) is statistically significant we must make further
assumptions about the data (conditional upon the fitted model) or the
model errors (residuals);
\(\varepsilon_t \stackrel{iid}{\sim} \mathcal{N}(0, \sigma^2)\). This
notation indicates that the residuals \(\varepsilon_t\) are
\emph{independent} and \emph{identically distributed} Gaussian random
variables with mean equal to \(0\) and constant variance \(\sigma^2\).
In the time series setting, the assumption of independence of model
residuals is often violated.

The linear model described above is quite restrictive in terms of the
types of trend it can fit; essentially linear increasing or decreasing
trends, or, trivially, a null trend of no change. This model can be
extended to allow for non-linear trends by making \(y_t\) depend on
polynomials of \(x_t\), for example

@@

where polynomials of \(x_t\) up to order \(P\) are used. This model
allows for more complex trends but it remains a fully parametric model
and suffers from several problems, especially the behaviour of the
fitted trend at the start and end of the observed series.

@@

Can we do better than these polynomial fits? In the remainder, I hope to
demonstrate that the answer to that question is emphatically ``yes''!
Below I describe a coherent and consistent approach to modelling
palaeoenvironmental time series using generalized additive models that
builds upon the linear regression framework.

@@

where the linear effect of time (the \(\beta_1 x_t\) part) has been
replaced by a smooth function of time, \(f(x_t)\). The immediate
advantage of the GAM is that we are no longer restricted to the shapes
of trends that can be fitted via global polynomial functions such as
\eqref{eq:polynomial-model}. Instead, the shape of the fitted trend will
be estimated from the data itself.

The linear model is a special case of a broader class, known as the
generalized linear model (GLM; McCullagh and Nelder, 1989). The GLM
provides a common framework for modelling a wide range of types of data,
such as count, proportions, or binary (presence/absence) data, that are
not conditionally distributed Gaussian. GLMs are, like the linear model,
parametric in nature; the types of trends that we can fit using a GLM
are the linear or polynomial models. GAMs extend the GLM by relaxing
this parametric assumption; in a GAM some, or all, of the parametric
terms, the \(\beta_p\), are replace by smooth functions \(f_j\) of the
covariates \(x_j\). For completeness then, we can write
\eqref{eq:additive-model} as a GLM/GAM

@@

where \(\mu_t\) is the expected value (e.g.~the mean count or the
probability of occurrence) of the random variable \(Y_t\)
(\(\mu_t \equiv \mathbb{E}(Y_t)\)) of which we have observations
\(y_t\). \(g\) is the link function, an invertible, monotonic function,
such as the natural logarithm, and \(g^{-1}\) is its inverse. The link
function maps values from the response scale on to the scale of the
linear predictor, whilst the inverse of the link function provides the
reverse mapping. For example, count data are strictly non-negative
integer values and are commonly modelled as a Poisson GLM/GAM using the
natural log link function. On the log scale, the response can take any
real value between \(-\infty\) and \(+\infty\), and it is on this scale
that model fitting actually occurs (i.e.~using equation
\eqref{eq:gam-linear-pred}). However we need to map these unbounded
values back on to the non-negative response scale. The inverse of the
log link function, the exponential function, achieves this and maps
values to the interval 0--\(\infty\) (equation \eqref{eq:gam-inverse}).

@@

It is clear from the plots of the data (Figure~\ref{fig:data-figure})
that we require the fitted trends for the Small Water
δ\textsuperscript{15}N and Braya-Sø \uk{} time series to be non-linear
functions, but it is less clear how to specify the actual shape require.
Ideally, we'd like to learn the shape of the trend from the data
themselves. We will refer to these non-linear functions as \emph{smooth
functions}, or just \emph{smooths} for short, and we will denote a
smooth using \(f(x_t)\). Further, we would like to represent the smooths
in a way that \eqref{eq:gam} is represented parametrically so that it
can be estimate within the well-studied GLM framework. This is achieved
by representing the smooth using a \emph{basis}. A basis is a set of
functions that collectively span a space of smooths that, we hope,
contains the true \(f(x_t)\) or a close approximation to it. The
functions in the basis are known as \emph{basis functions}, and arise
from a \emph{basis expansion} of a covariate. Writing \(b_j(x_t)\) as
the \(j\)th basis function of \(x_t\), the smooth \(f(x_t)\) can be
represented as a weighted sum of basis functions

@@

where \(\beta_j\) is the weight applied to the \(j\)th basis function.

The polynomial model is an example of a statistical model that uses a
basis expansion. For the cubic polynomial (\(P = 3\)) fit shown in
Figure \ref{fig:polynomial-example-plot} there are in fact 4 basis
functions: \(b_1(x_t) = x_t^0 = 1\), \(b_2(x_t) = x_t\),
\(b_3(x_t) = x^2_t\), and \(b_4(x_t) = x_t^3\). Note that \(b_1(x_t)\)
is constant and is linked to the model intercept, \(\beta_0\), in the
linear model \eqref{eq:polynomial-model}, and further, that the basis
function weights are the estimated coefficients in the model, the
\(\beta_j\).

@@

A cubic spline is a smooth curve comprised of sections of cubic
polynomials, where the sections are joined together at some specified
locations --- known as \emph{knots} --- in such a way that at the joins,
the two sections of cubic polynomial that meet have the same value as
well as the same first and second derivative. These properties mean that
the sections join smoothly and differentiably at the knots (Wood, 2017,
5.3.1).

@@

To estimate a model using this basis expansion each basis function forms
a column in the model matrix \textbf{X} and the weights \(\beta_j\) can
be found using least squares regression (assuming a Gaussian response).
Note that in order to estimate a coefficient for each basis function the
model has to be fitted without an intercept term. In practice we would
include an intercept term in the model and therefore the basis functions
are modified via an identifiability constraint (@ Wood, 2017). This has
the effect of making the basis orthogonal to the intercept but results
in more complicated basis functions than those shown in in Figure
\ref{fig:basis-function-example-plot}a.

@@

Cubic regression splines, as well as many other types of spline, require
the analyst to choose the number and location of the knots that
parametrise the basis. Thin plate regression splines (TPRS) remove this
element of subjectivity when fitting GAMs. Thin plate splines were
introduced by Duchon (1977) and, as well as solving the knot selection
problem, have several additional attractive properties in terms of
optimality and their ability to estimate a smooth function of two or
more variables, leading to smooth interactions between covariates.
However, thin plate splines have one key disadvantage over CRS; thin
plate splines have as many unknown parameters as there are unique
combinations of covariate values in a data set (Wood, 2017, 5.5.1). It
is unlikely that any real data problem would involve functions of such
complexity that they require as many basis functions as data. It is much
more likely that the true functions that we attempt to estimate are far
simpler than the set of functions representable by 1 basis function per
unique data value. From a practical point of view, it is also highly
inefficient to carry around all these basis functions whilst model
fitting, and the available computational resources would become quickly
exhausted for large time series with many observations.

To address this issue low rank thin plate regression splines (TPRS) have
been suggested which truncate the space of the thin plate spline basis
to some lower number of basis functions whilst preserving much of the
advantage of the original basis as an optimally-fitting spline (Wood,
2003). A rank 7 TPRS basis (i.e.~one containing 7 basis functions) is
shown in Figure \ref{fig:basis-function-example-plot}c for the Small
Water time series. The truncation is achieved by performing an
eigen-decomposition of the basis functions and retaining the
eigenvectors associated with the \(k\) largest eigenvalues. This is
similar to the way principal components analysis decomposes a data set
into axes of variation (eigenvectors) in decreasing order of variance
explained. The truncated basis can preserve much of the space of
functions spanned by the original basis but at the cost of using far
fewer basis functions (Wood, 2003, 2017, 5.5.1). Note the horizontal
TPRS basis function (at δ\textsuperscript{15}N = 1) in Figure
\ref{fig:basis-function-example-plot}c; this basis function is
confounded with the intercept term and, after the application of
identifiability constraints, ends up being removed from the set of basis
functions used to fit the model.

The truncation suggested by Wood (2003) is not without cost; the
eigen-decomposition and related steps can be relatively costly for large
data sets. For data sets of similar size to the two examples used here,
the additional computational effort required to set up the TPRS basis
over the CRS basis will not be noticeable. For highly resolved series
containing more than \textasciitilde{}1000 observations the truncation
may be costly computationally. In such instances, little is lost by
moving to the CRS basis with the same number of knots as the rank of the
desired TPRS, with the benefit of considerably reduced set up time for
the basis.

@@

In practice, the size of the basis is an upper limit on the expected
complexity of the trend, and a simple test to check if the basis used
was sufficiently large (Pya and Wood, 2016). This test is available via
the \texttt{gam.check()} function in \emph{mgcv} for example, which
essentially looks at whether there is any additional nonlinearity or
structure in the residuals that can be explained by a further smooth of
\(x_t\). Should a smooth term in the fitted model fail this test the
model can be refitted using a larger basis expansion, say by doubling
the value of \emph{k} (the rank) used to fit the original. Note also
that a smooth might fail this test whilst using fewer effective degrees
of freedom than the maximum possible for the dimension of basis used.
This may happen when the true function lies at the upper limit of the
set of functions encompassed by the size of basis used. Additionally, a
basis of size \(2k\) encompasses a richer space of functions of a given
complexity than a basis of size \(k\) (Wood, 2017); increasing the basis
dimension used to fit the model may unlock this additional function
space resulting in a better fitting model whilst using a similar number
of effective degrees of freedom.

@@

Having identified low rank regression splines as a useful way to
represent \(f\), we next need a way to decide how wiggly the fitted
trend should be. A backwards elimination approach to sequentially remove
knots or basis functions might seem appropriate, however such an
approach would likely fail as the resulting sequence of models would not
be strictly nested, precluding many forms of statistical comparison
(Wood, 2017). Alternatively, we could keep the basis dimension at a
fixed size but guard against fitting very complex models through the use
of a wiggliness penalty.

@@

The trend in δ\textsuperscript{15}N values is clearly non-linear but it
would be difficult to suggest a suitable polynomial model that would
allow for periods of relatively no change in δ\textsuperscript{15}N as
well as rapid change. Instead, a GAM is ideally suited to modelling such
trends; the data suggest a smoothly varying change in
δ\textsuperscript{15}N between 1925 and 1975. It is reasonable to expect
some autocorrelation in the model errors about the fitted trend.
Therefore I fitted the following GAM to the δ\textsuperscript{15}N time
series.

@@

Fitting this model as a standard GAM with REML smoothness selection
resulted in the same fitted trend as the GAM with CAR(1) errors (not
shown), whilst using GCV smoothness selection resulted in a much more
satisfactory fitted trend. There are two potential problems with the
GCV-selected trend: i) GCV is sensitive to the profile of the GCV score
and has been shown to under smooth data in situations where the profile
is flat around the minimum GCV score, and ii) the model fitted assumes
that the observations are independent, an assumption that is certainly
violated in the \uk{} time series.

To investigate the first issue, the GCV and REML scores for an
increasing sequence of values of the smoothness parameter (\(\lambda\))
were evaluated for the standard GAM (equation \eqref{eq:gam}) fit to the
\uk{} time series. The resulting profiles are shown in Figure
\ref{fig:trace-smoothness-parameters}, with the optimal value of the
parameter shown by the vertical line. The GCV score profile suggests
that the potential for under smoothing identified by Reiss and Ogden
(2009) is unlikely to apply here as there is a well-defined minimum in
profile.
